---
title: "Drafting Visualizations"
format: html
---

```{r}
#| message: FALSE
#| warning: FALSE
# Import libraries
library(tidyverse)
library(here)
library(ggplot2)
library(plotly)
library(showtext)

# Read in data 
bloom <- read_csv(here::here("data", "bloom_clean.csv"))

# Add fonts
font_add_google(name = "Rubik", family = "rubik") # Headings and titles
font_add_google(name = "Karla", family = "karla") # Subtitles, axes, and captions
showtext_auto()

# Create color palette 
 bloom_palette <- c("Wildflower center green" = "#60721C", "White" = "#F9FAFA", "floral1" = "#EA71D5", "floral2" = "#F0A0C6")
 
 
# Create list of common names to pull from as-needed
 common_names <- tribble(
  ~scientific_name,                       ~common_name,
  "Salvia farinacea",                     "Mealy blue sage",
  "Pavonia lasiopetala",                  "Rock rose",
  "Berlandiera lyrata",                   "Chocolate daisy",
  "Malvaviscus arboreus var. drummondii", "Turk's cap",
  "Conoclinium dissectum",                "Gregg's mistflower",
  "Salvia greggii",                       "Autumn sage",
  "Echinacea purpurea",                   "Purple coneflower",
  "Lupinus texensis",                     "Bluebonnet",
  "Glandularia bipinnatifida",            "Prairie verbena",
  "Lantana urticoides",                   "Texas lantana",
  "Ilex decidua",                         "Possumhaw",
  "Gelsemium sempervirens",               "Carolina Jessamine",
  "Liatris punctata var. mucronata",      "Texas blazing star",
  "Senna lindheimeriana",                 "Lindheimer's senna",
  "Helianthus maximiliani",               "Maximilian sunflower",
  "Ilex vomitoria",                       "Yaupon holly",
  "Viguiera stenoloba",                   "Skeleton-leaf goldeneye"
)
```

### Step 1: Collect statistics for use in header

#### Total species count 
How many unique species were observed over the full time period?
```{r}
# Count unique scientific names
totalspecies <- bloom %>% 
  summarise(unique_count = n_distinct(scientific_name))
```
-> 474 total species.

#### Total observations
How many plants were listed total over the full time period?
```{r}
# Number of rows where scientific name is not NA
nrow(bloom %>% filter(!is.na(scientific_name)))
```

#### Number of plants observed once
```{r}
# Unique observations
unique <- bloom %>%
  filter(!is.na(scientific_name)) %>%
  mutate(year = year(date)) %>%
  count(scientific_name) %>%
  filter(n == 1) %>%
  left_join(bloom %>% mutate(year = year(date)), by = "scientific_name") %>%
  select(scientific_name, year, n)

nrow(unique)
```

### Step 2: Bar plot of top species

#### What were the top 10 most commonly noted plants over full time period?
```{r}
# Top 10 most commonly noted plants over full data
top10 <- bloom %>%
  filter(!is.na(scientific_name)) %>%
  count(scientific_name) %>%
  slice_max(n, n = 10, with_ties = FALSE)

# Add common names for plotting
top10 <- top10 %>% 
  left_join(common_names, by = "scientific_name")
```

```{r}
# Create bar plot
p1 <- ggplot(top10, aes(x = common_name, y = n)) +
  geom_col(fill = bloom_palette[4]) +
  labs(title = "What were the top 10 most commonly listed plants?",
       subtitle = "These 10 plants were noted by volunteers over 100 times between the years of 2011 and 2026,\nmeaning they are very commonly in bloom, growing, making seeds, or otherwise noticable!",
       x = NULL,
       y = "Count") +
  theme_minimal(base_size = 12) +
  theme(
    axis.text.x = element_text(angle = 35, hjust = 1),
    plot.title = element_text(family = "rubik", face = "bold", size = rel(1.9), margin = margin(b = 6)),
    plot.subtitle = element_text(family = "karla", size = rel(1.2)))

p1

ggsave(here("figures", "top10bar.pdf"), width = 10, height = 8, dpi = 150, bg = "white")
```

### Step 3: Phenology heat maps

#### Are there any changing seasonal patterns in the most common species?

```{r}

top4 <- top10 %>% slice_max(n, n = 4, with_ties = FALSE)

grid_data <- bloom %>%
  filter(scientific_name %in% top4$scientific_name) %>%
  mutate(year = year(date), month = month(date, label = TRUE)) %>%
  count(scientific_name, year, month) %>%
  complete(scientific_name, year, month, fill = list(n = 0))

p <- ggplot(grid_data, aes(x = month, y = factor(year), fill = n)) +
  geom_tile(color = "white") +
  scale_fill_gradient(low = "grey90", high = "darkgreen") +
  facet_wrap(~scientific_name) +
  labs(x = "Month", y = "Year", fill = "Count")

ggplotly(p)
```


```{r}
bloom_shift <- bloom %>%
  filter(scientific_name %in% top4$scientific_name) %>%
  mutate(year = year(date)) %>%
  filter(!year %in% c(2011, 2026)) %>%
  mutate(day_of_year = yday(date)) %>%
  group_by(scientific_name, year) %>%
  summarise(first_seen = min(day_of_year), last_seen = max(day_of_year), .groups = "drop")

p <- ggplot(bloom_shift, aes(x = year)) +
  geom_line(aes(y = first_seen, linetype = "First observed")) +
  geom_line(aes(y = last_seen, linetype = "Last observed")) +
  facet_wrap(~scientific_name) +
  labs(x = "Year", y = "Day of year", linetype = "")

ggplotly(p)
```

### Step 4: Bar plot of monthly plants

#### What were the most commonly observed plants in each month of the year?

```{r}
# Most observed plants each month (2011-2026)
top_by_month <- bloom %>%
  filter(!is.na(scientific_name)) %>%
  mutate(month = month(date, label = TRUE)) %>%
  count(month, scientific_name) %>%
  group_by(month) %>%
  slice_max(n, n = 1, with_ties = TRUE) %>%
  ungroup()

top_by_month <- top_by_month %>% 
  left_join(common_names, by = "scientific_name")
```

