---
title: "Drafting Visualizations"
format: html
---

```{r}
#| message: FALSE
#| warning: FALSE
# Import libraries
library(tidyverse)
library(here)
library(ggplot2)
library(plotly)
library(showtext)
library(ggtext)

# Read in data 
bloom <- read_csv(here::here("data", "bloom_clean.csv"))

# Add fonts
font_add_google(name = "Rubik", family = "rubik") # Headings and titles
font_add_google(name = "Karla", family = "karla") # Subtitles, axes, and captions
showtext_auto()

# Create color palette 
 bloom_palette <- c("Wildflower center green" = "#60721C", "White" = "#F9FAFA", "floral1" = "#EA71D5", "floral2" = "#F0A0C6", "floral3" = "#5F7ECB")
 
 # Select species colors for plot- color matche from flower images
species_colors <- c(
  "Turk's cap" = "#FA3B25",
  "Texas lantana" = "#F9740A",
  "Rock rose" = "#F865AE",
  "Purple coneflower" = "#E49CEC",
  "Prairie verbena" = "#7D7ADA",
  "Mealy blue sage" = "#5677EE",
  "Gregg's mistflower" = "#BDB7DB",
  "Chocolate daisy" = "#D9C703",
  "Bluebonnet" = "#0B67BF",
  "Autumn sage" = "#DE03B5"
)
 
 
# Create list of common names to pull from as-needed
 common_names <- tribble(
  ~scientific_name,                       ~common_name,
  "Salvia farinacea",                     "Mealy blue sage",
  "Pavonia lasiopetala",                  "Rock rose",
  "Berlandiera lyrata",                   "Chocolate daisy",
  "Malvaviscus arboreus var. drummondii", "Turk's cap",
  "Conoclinium dissectum",                "Gregg's mistflower",
  "Salvia greggii",                       "Autumn sage",
  "Echinacea purpurea",                   "Purple coneflower",
  "Lupinus texensis",                     "Bluebonnet",
  "Glandularia bipinnatifida",            "Prairie verbena",
  "Lantana urticoides",                   "Texas lantana",
  "Ilex decidua",                         "Possumhaw",
  "Gelsemium sempervirens",               "Carolina Jessamine",
  "Liatris punctata var. mucronata",      "Texas blazing star",
  "Senna lindheimeriana",                 "Lindheimer's senna",
  "Helianthus maximiliani",               "Maximilian sunflower",
  "Ilex vomitoria",                       "Yaupon holly",
  "Viguiera stenoloba",                   "Skeleton-leaf goldeneye"
)
```

### Step 1: Collect statistics for use in header

#### Total species count 
How many unique species were observed over the full time period?
```{r}
# Count unique scientific names
totalspecies <- bloom %>% 
  summarise(unique_count = n_distinct(scientific_name))
```
-> 474 total species.

#### Total observations
How many plants were listed total over the full time period?
```{r}
# Number of rows where scientific name is not NA
nrow(bloom %>% filter(!is.na(scientific_name)))
```

#### Number of plants observed once
```{r}
# Unique observations
unique <- bloom %>%
  filter(!is.na(scientific_name)) %>%
  mutate(year = year(date)) %>%
  count(scientific_name) %>%
  filter(n == 1) %>%
  left_join(bloom %>% mutate(year = year(date)), by = "scientific_name") %>%
  select(scientific_name, year, n)

nrow(unique)
```

### Step 2: Bar plot of top species

#### What were the top 10 most commonly noted plants over full time period?
```{r}
# Top 10 most commonly noted plants over full data
top10 <- bloom %>%
  filter(!is.na(scientific_name)) %>%
  count(scientific_name) %>%
  slice_max(n, n = 10, with_ties = FALSE)

# Add common names for plotting
top10 <- top10 %>% 
  left_join(common_names, by = "scientific_name")
```

```{r}
# Create bar plot
p1 <- ggplot(top10, aes(x = fct_reorder(common_name, n, .desc = TRUE), y = n, fill = common_name)) +
  geom_col(alpha = 0.75) +
  scale_fill_manual(values = species_colors, guide = "none") +
  labs(x = NULL,
       y = NULL) +
  theme_minimal(base_size = 12) +
  theme(
    plot.title = element_text(family = "rubik", face = "bold", size = rel(1.9), margin = margin(b = 6)),
    plot.subtitle = element_text(family = "karla", size = rel(1.2)),
    axis.text.x = element_text(angle = 35, hjust = 1, size = 12.5, family = "karla"),
    axis.text.y = element_text(size = 12.5, family = "karla"))

p1

ggsave(here("figures", "top10bar.pdf"), width = 10, height = 8, dpi = 150, bg = "white", device = cairo_pdf)
```


### Step 3: Phenology heat maps

#### Are there any changing seasonal patterns in the most common species?

```{r}
# Filter data to top 4 most common species
heatmap_data1 <- bloom %>% 
  filter(scientific_name %in% c("Salvia farinacea")) %>% 
  mutate(year = year(date), month = month(date, label = TRUE)) %>% 
  count(scientific_name, year, month) %>%
  complete(scientific_name, year, month, fill = list(n = 0))

#.........................create heatmap.........................
p2 <- ggplot(heatmap_data1, aes(x = month, y = year, fill = n)) +
  geom_tile(color = "white", linewidth = 0.5) +
  labs(fill = "Observation count",
       title = "**Mealy blue sage** <span style='font-size:14pt'>*(Salvia farinacea)*</span>") + 
  scale_fill_gradientn(colors = c("#faeef0", bloom_palette[5])) +
  scale_y_continuous(breaks = seq(2012, 2026, by = 2)) +
  guides(fill = guide_colorbar(barwidth = 15, barheight = 0.75, title.position = "top")) +
  theme_minimal() +
  theme(
    legend.position = "bottom",
    legend.title = element_text(family = "karla", hjust = 0.5),
    axis.title = element_blank(),
    panel.grid = element_blank(),
    axis.text.x = element_text(family = "karla"),
    plot.title = element_markdown(family = "rubik", size = rel(1.8), margin = margin(b = 6)),
    plot.subtitle = element_text(family = "karla", size = rel(1.2)),
  )

p2

ggsave(here("figures", "heatmap1.pdf"), width = 10, height = 10, dpi = 150, bg = "white")
```

```{r}
heatmap_data2 <- bloom %>% 
  filter(scientific_name %in% c("Pavonia lasiopetala")) %>% 
  mutate(year = year(date), month = month(date, label = TRUE)) %>% 
  count(scientific_name, year, month) %>%
  complete(scientific_name, year, month, fill = list(n = 0))

#.........................create heatmap.........................
p3 <- ggplot(heatmap_data2, aes(x = month, y = year, fill = n)) +
  geom_tile(color = "white", linewidth = 0.5) +
  labs(fill = "Observation count") + 
  scale_fill_gradientn(colors = c("#f5e1f0", "#F89BEA", "#FC47B0")) +
  scale_y_continuous(breaks = seq(2012, 2026, by = 2)) +
  guides(fill = guide_colorbar(barwidth = 15, barheight = 0.75, title.position = "top")) +
  scale_x_discrete(expand = c(0, 0)) +
  theme_minimal() +
  theme(
    legend.position = "bottom",
    legend.title = element_text(family = "karla", hjust = 0.5, size = 24),
    legend.text = element_text(family = "karla", size = 16),
    axis.title = element_blank(),
    panel.grid = element_blank(),
    axis.text.x = element_text(family = "karla", size = 20, margin = margin(t = -10)),
    axis.text.y = element_text(family = "karla", size = 20)
  )

p3

ggsave(here("figures", "rockrose.pdf"), width = 10, height = 10, dpi = 150, bg = "white")
```


```{r}
heatmap_data3 <- bloom %>% 
  filter(scientific_name %in% c("Berlandiera lyrata")) %>% 
  mutate(year = year(date), month = month(date, label = TRUE)) %>% 
  count(scientific_name, year, month) %>%
  complete(scientific_name, year, month, fill = list(n = 0))

#.........................create heatmap.........................
p4 <- ggplot(heatmap_data3, aes(x = month, y = year, fill = n)) +
  geom_tile() +
  labs(fill = "Observation count",
       title = "**Chocolate daisy** <span style='font-size:14pt'>*(Berlandiera lyrata)*</span>") + 
  scale_fill_gradientn(colors = c("#faeef0", bloom_palette[4], bloom_palette[3])) +
  scale_y_continuous(breaks = seq(2012, 2026, by = 2)) +
  guides(fill = guide_colorbar(barwidth = 15, barheight = 0.75, title.position = "top")) +
  theme_minimal() +
  theme(
    legend.position = "bottom",
    legend.title = element_text(family = "karla", hjust = 0.5),
    axis.title = element_blank(),
    panel.grid = element_blank(),
    axis.text.x = element_text(family = "karla"),
    plot.title = element_markdown(family = "rubik", size = rel(1.8), margin = margin(b = 6)),
    plot.subtitle = element_text(family = "karla", size = rel(1.2)),
  )

p4

ggsave(here("figures", "heatmap3.pdf"), width = 10, height = 8, dpi = 150, bg = "white")
```
```{r}
heatmap_data4 <- bloom %>% 
  filter(scientific_name %in% c("Malvaviscus arboreus var. drummondii")) %>% 
  mutate(year = year(date), month = month(date, label = TRUE)) %>% 
  count(scientific_name, year, month) %>%
  complete(scientific_name, year, month, fill = list(n = 0))

#.........................create heatmap.........................
p5 <- ggplot(heatmap_data4, aes(x = month, y = year, fill = n)) +
  geom_tile() +
  labs(fill = "Observation count",
       title = "**Turk's cap** <span style='font-size:14pt'>*(Malvaviscus arboreus var. drummondii)*</span>") + 
  scale_fill_gradientn(colors = c("#faeef0", bloom_palette[4], bloom_palette[3])) +
  scale_y_continuous(breaks = seq(2012, 2026, by = 2)) +
  guides(fill = guide_colorbar(barwidth = 15, barheight = 0.75, title.position = "top")) +
  theme_minimal() +
  theme(
    legend.position = "bottom",
    legend.title = element_text(family = "karla", hjust = 0.5),
    axis.title = element_blank(),
    panel.grid = element_blank(),
    axis.text.x = element_text(family = "karla"),
    plot.title = element_markdown(family = "rubik", size = rel(1.8), margin = margin(b = 6)),
    plot.subtitle = element_text(family = "karla", size = rel(1.2)),
  )

p5

ggsave(here("figures", "heatmap4.pdf"), width = 10, height = 8, dpi = 150, bg = "white")
```


```{r}
heatmap_data5 <- bloom %>% 
  filter(scientific_name %in% c("Lantana urticoides")) %>% 
  mutate(year = year(date), month = month(date, label = TRUE)) %>% 
  count(scientific_name, year, month) %>%
  complete(scientific_name, year, month, fill = list(n = 0))

#.........................create heatmap.........................
p6 <- ggplot(heatmap_data5, aes(x = month, y = year, fill = n)) +
  geom_tile(color = "white", linewidth = 0.5) +
  labs(fill = "Observation count",
       title = "**Texas Lantana** <span style='font-size:14pt'>*(Pavonia lasiopetala)*</span>") + 
  scale_fill_gradientn(colors = c("#f7fbe8","#E2C400", "#F3A404", "#F9740A")) +
  scale_y_continuous(breaks = seq(2012, 2026, by = 2)) +
  guides(fill = guide_colorbar(barwidth = 15, barheight = 0.75, title.position = "top")) +
  theme_minimal() +
  theme(
    legend.position = "bottom",
    legend.title = element_text(family = "karla", hjust = 0.5),
    axis.title = element_blank(),
    panel.grid = element_blank(),
    axis.text.x = element_text(family = "karla"),
    plot.title = element_markdown(family = "rubik", size = rel(1.8), margin = margin(b = 6)),
    plot.subtitle = element_text(family = "karla", size = rel(1.2)),
  )

p6

ggsave(here("figures", "heatmap5.pdf"), width = 10, height = 10, dpi = 150, bg = "white")
```

### Step 4: Bloom calendar heatmap

#### When were the top 10 most common plants blooming throughout the year?

```{r}
# Count of observations of each species each month
species_month_counts <- bloom %>%
  mutate(month = month(date, label = TRUE)) %>%
  count(scientific_name, month)

# Fill all missing months with 0
species_month_counts <- species_month_counts %>%
  complete(scientific_name, month, fill = list(n = 0))

# Filter to just top 10 most common species
species_month_counts <- species_month_counts %>%
  filter(scientific_name %in% top10$scientific_name) %>%
  left_join(common_names, by = "scientific_name")

```

```{r}
# Test plot
ggplot(species_month_counts, aes(x = month, y = common_name, fill = n)) +
  geom_tile(color = "white", linewidth = 0.5) +
  scale_fill_gradient(low = "white", high = "darkgreen") +
  theme_minimal() +
  labs(x = NULL, y = NULL, fill = "Observations")
```

```{r}
# Design heatmap
species_month_counts <- species_month_counts %>%
  mutate(common_name = fct_reorder(common_name, n, .fun = sum, .desc = TRUE)) %>% 
  group_by(common_name) %>%
  mutate(
    tile_color = {
      ramp <- colorRampPalette(c("#899E2E", species_colors[as.character(common_name[1])]))
      ramp(100)[as.integer(scales::rescale(n, to = c(1, 100)))]
    }
  ) %>%
  ungroup()

ggplot(species_month_counts, aes(x = as.numeric(month), y = 1, fill = tile_color)) +
  geom_raster(interpolate = TRUE) +
  scale_fill_identity() +
  facet_wrap(~common_name, ncol = 1, strip.position = "left") +
  scale_x_continuous(breaks = 1:12, labels = month.abb, expand = c(0, 0)) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(size = 16, family = "karla"),
    axis.text.y = element_blank(),
    axis.ticks.y = element_blank(),
    strip.text.y.left = element_text(angle = 0, hjust = 1, size = 13, family = "karla"),
    panel.spacing = unit(0, "lines")
  ) +
  labs(x = NULL, y = NULL)

 ggsave(here("figures", "seasonalheatmap.pdf"), width = 19, height = 6, dpi = 300)
```


### Step 5: Plot of monthly plants

#### What were the most commonly observed plants in each month of the year?

```{r}
# Most observed plants each month (2011-2026)
top_by_month <- bloom %>%
  filter(!is.na(scientific_name)) %>%
  mutate(month = month(date, label = TRUE)) %>%
  count(month, scientific_name) %>%
  group_by(month) %>%
  slice_max(n, n = 1, with_ties = TRUE) %>%
  ungroup()

top_by_month <- top_by_month %>% 
  left_join(common_names, by = "scientific_name")
```

